---
title: "Introduction to data visualisation with `ggplot2`"
author: "Jonas Hagenberg"
institute: "EPN autumn school"
date: "10.09.2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(ggplot2)
library(magrittr)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1a5f96",
  secondary_color = "#03A696",
  code_inline_background_color = "#f8f8f8"
)
```

# Overview

- read in data
- bring it into the correct form
- plot it with `ggplot2`

---

# Data read in
```{r}
cytokine_data <- read.table(
  file = "03_data/example_cytokines.csv",
  header = TRUE,
  sep = ";"
)
head(cytokine_data)
```

---

# `ggplot2` - grammar of data

---

# Overview `ggplot2`
```{r, eval = FALSE}
library(ggplot2)
ggplot(
  data = your_dataset,
  mapping = aes(x = variable)
) +
  geom_*
```

For better clarity it can also be written as:
```{r, eval = FALSE}
library(magrittr)
dataset %>% 
  ggplot() +
  aes(x = variable) +
  geom_*
```

---

# Simple Boxplot
```{r}
cytokine_data %>% 
  ggplot() +
  aes(y = TNF_alpha) +
  geom_boxplot()
```

---

# Tidy data
So far: every row is one subject, several cytokines per row.
**Tidy data:**
- Every column is a variable.
- Every row is an observation.
- Every cell is a single value.

We only want **one observation per row**.

---

# From wide to long format
```{r, message = FALSE}
library(tidyr)
cytokine_data_long <- pivot_longer(
  data = cytokine_data,
  cols = TNF_alpha:IL_17A,
  names_to = "assay",
  values_to = "concentration"
)
head(cytokine_data_long)
```

---

# Several boxplots
Now we can use the new variables in `ggplot2`:
```{r}
cytokine_data_long %>% 
  ggplot() +
  aes(
    x = assay,
    y = concentration,
    fill = assay
  ) +
  geom_boxplot()
```

