---
title: "Introduction to data visualisation with `ggplot2`"
author: "Jonas Hagenberg"
institute: "EPN autumn school"
date: "09.09.2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(ggplot2)
library(magrittr)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1a5f96",
  secondary_color = "#03A696",
  code_inline_background_color = "#f8f8f8"
)
```

# Overview

- read in data
- bring it into the correct form
- plot it with `ggplot2`

---

# Boxplot
```{r, echo = FALSE}
set.seed(4)
bp_data <- data.frame(test = rnorm(20))
anno_data <- data.frame(
  text = c("outlying point", "25% quantile - 1.5 * IQR", "25% quantile", "median",
           "75% quantile", "75% quantile + 1.5 * IQR"),
  pos = c(-1.281247, -0.5424926, -0.05894495, 0.1928908, 0.7397427, 1.89654)
)
ggplot(
  data = bp_data,
  mapping = aes(x = 0.5, y = test)
) +
  geom_boxplot() +
  geom_point(mapping = aes(x = 0, y = test)) +
  geom_text(
    data = anno_data,
    mapping = aes(x = 1, y = pos, label = text),
    hjust = "left",
    size = 8
    ) +
  theme_minimal(base_size = 22) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  ) +
  xlab(NULL) +
  ylab(NULL) +
  coord_cartesian(xlim = c(0, 3))
```


# Data read in
```{r}
cytokine_data <- read.table(
  file = "03_data/example_cytokines.csv",
  header = TRUE,
  sep = ";"
)
head(cytokine_data)
```

---

# `ggplot2` - grammar of data

describe your plot step by step:
- which data set to use
- map variables to aesthetics (e.g. x position, y position, fill colour, shape, ...)
- which geometric objects to choose (e.g. points, lines, ...)
- which scale to use (untransformed scale, log scale, ...)
- further options how your plot should look like (colours, themes, grids, ...)

---

# Overview `ggplot2`
```{r, eval = FALSE}
library(ggplot2)
ggplot(
  data = your_dataset,
  mapping = aes(x = variable)
) +
  geom_*
  ```

---

# Simple Boxplot
.pull-left[
```{r, eval = FALSE}
ggplot(
  data = cytokine_data,
  mapping = aes(y = TNF_alpha)
) +
  geom_boxplot()
```
]
.pull-right[
```{r, echo = FALSE}
ggplot(
  data = cytokine_data,
  mapping = aes(y = TNF_alpha)
) +
  geom_boxplot()
```
]


---

# Tidy data
So far: every row is one subject, several cytokines per row.

**Tidy data:**
- Every column is a variable.
- Every row is an observation.
- Every cell is a single value.

We only want **one observation per row**.

---

# From wide to long format
```{r, message = FALSE}
library(tidyr)
cytokine_data_long <- pivot_longer(
  data = cytokine_data,
  cols = TNF_alpha:IL_17A,
  names_to = "assay",
  values_to = "concentration"
)
head(cytokine_data_long)
```

---

# Several boxplots
.pull-left[
Now we can use the new variables in `ggplot2`:
```{r, eval = FALSE}
ggplot(
  data = cytokine_data_long,
  mapping = aes(
    x = assay,
    y = concentration,
    fill = assay
  )
) +
  geom_boxplot()
```
]
.pull-right[
```{r, echo = FALSE}
ggplot(
  data = cytokine_data_long,
  mapping = aes(
    x = assay,
    y = concentration,
    fill = assay
  )
) +
  geom_boxplot()
```
]

---

# Where to find help

- documentation: https://ggplot2.tidyverse.org/index.html
- https://community.rstudio.com/
- http://www.sthda.com/english/wiki/ggplot2-essentials
- https://r-graph-gallery.com/ggplot2-package.html
- https://stackoverflow.com/questions/tagged/r

---

# Alternative writing style
For better clarity one can also write:
```{r, eval = FALSE}
library(magrittr)
dataset %>% 
  ggplot() +
  aes(x = variable) +
  geom_*
```
  
  